Associations between Chlorophyll \emph{a} and various Microcystin-LR Health Advisory Concentrations

### 2015-10-01 manuscript/manuscript.tex
jeff read through done.  Send to betty.  Plus a render

Over the last decade or so numerous events and legislative activities
have raised the public awareness of harmful algal blooms (Rinta-Kanto et
al. 2009, HABHRCA 2014, Jetoo et al. 2015). Recently, the US
Environmental Protection Agency (USEPA) has released suggested
microcystin-LR (one of the more common toxins) concentrations that would
trigger health advisories (McElhiney and Lawton 2005, Zurawell et al.
2005, USEPA 2015). Additionally, the World Health Organization (WHO) has
had proposed advisory levels for drinking water and a range of
recreational risk levels (Chorus and Bartram 1999, (WHO) and others
2003). While these levels and associated advisories are likely to help
mitigate the impacts from harmful algal blooms, they are not without
complications.

One of these complications is that they rely on available measurements
of microcystin-LR. While laboratory testing remains the gold standard
for quantifying microcystin-LY concentrations in water samples, several
field test kits have been developed. Even though field tests provide a
much needed means for rapid assessment, they are infrequently available
to lake managers. These kits are moderately expensive (approximately
\$150-\$200 depending on specific kit) with a limited shelf life
(typically one year) (James et al. 2011, Aranda-Rodriguez et al. 2015).
Additionally, each technique requires nuanced understanding of the
detection method (e.g., limit of detection, specific microcystin
variants being measured, and sampling protocol). Fortunately,
microcystin-LR has been shown to be associated with several other, more
commonly measured and understood components of water quality.

Chlorophyll \emph{a} is a very commonly measured components of water
quality that is also known to be associated with Microsystin-LR
concentrations {[}Lee et al. (2000); Paerl and Otten (2013);
heisler2008eutrophication{]}. Additionally there are many rapid
measurements for assessing chlorophyll \emph{a} levels \emph{in situ}.
For instance, there are small or hand held flourometers that measure
chlorohpyll \emph{a}. Given these facts, it might be possible to
identify chlorophyll \emph{a} concentrations that would be associated
with the various Microcystin-LR health advisory levels. Identifying
these associations would provide another reliable tool for water
resource managers to use to help manage the threat to public health
posed by cHABs and would be especially useful in the absence of
microcystin-LR concentrations. Thus, the goal of this paper is to utlize
the National Lakes Assessment data and identify chlorophyll \emph{a}
concentrations that are associated with higher probabilities of
exceeding several microcystin-LR health advisory concentrations (USEPA
2009). So that others may repeat or adjust this analysis, the data,
code, and this manuscript are freely available via
\href{https://github.com/USAPE/microcystinchla}{\url{https://github.com/USAPE/microcystinchla}}.

@@

A detailed discussion of CPA is beyond the scope of this paper, but see
Paul et al. (2005) and Hollister et al. (2008) for details. For this
analysis, we used CPA to examine how the conditional probability of
exceeding one of the health advisory changes as chlorophyll \emph{a}
increases in a lake. The 95\% confidence intervals were calculated from
1000 bootstrapped samples. To identify chlorophyll \emph{a}
concentrations of concern we used a 50\% conditional probability of
exceeding each health advisory level and extracted the minimum
chlorophyll \emph{a} concentration that was associated with the upper
confidence level being 50\% or greater. As both microcystin-LR and
chlorophyll \emph{a} values were both highely skewed right, a log base
10 transformation was used. Additional details of the specific
implementation are available at
\url{https://github.com/USEPA/microcystinchla}.

@@

In the 2007 NLA, microcystin-LR concentrations ranged from 0.05 to 225.
Microcystin-LR concentrations of 0.05 \(\mu\)g/L represent the detection
limits. Any value greater than that indicates the presence of
microcystin-LR. Of those lakes with microcystin, the median
concentration was 0.51 and the mean was 0.51. Lastly, of all lakes
sampled, 21\% of lakes exceeded the U.S. EPA childrens drinking water
standard, 8.8\% of lakes exceeded the U.S. EPA adult drinking water
standard,11.7\% of lakes exceeded the WHO drinking water standard,and
7.3\% of lakes exceeded the WHO recreational standard. For chlorophyll
\emph{a}, the range was 0.07 to 936. All lakes had reported chlorophyll
\emph{a} concentrations that exceeded detection limits. The median
concentration was 7.79 and the mean was 29.6301946. The associations
between chlorophyll \emph{a} and the upper confidence interval with a
conditional probability of 50\% were 22.25, 65.6, 84.96, and 105.84 for
0.3, 1.0, 1.6 and 2.0 \(\mu\)g/L, respectively (Table
\ref{tab:mc_chla_table} \& Figure \ref{fig:multi_cp_plot}).

@@

There are numerous possible uses for the chlorophyll \emph{a} and
microcystin-LR advisory cut-off values. First, in the absence of
microcystin-LR measurements, exceedence of the chlorophyll \emph{a}
concentrations could be a trigger for further actions. Given that there
is uncertainity around these chlorophyll \emph{a} cutoffs the best case
scenario would be to monitor for chlorophyll and in the event of
exceeding a target concentration take water samples and have those
samples tested in a lab for microcystin-LR. A second potential use is to
identify possible bloom events from historical data. As harmful algal
blooms are made up of many species and have various mechanism
responsible for adverse impacts (e.g.~toxins, hypoxia, odors), there is
no single definition of a bloom. For cHABs one approach has been to
identify an increase over a baseline concentration of phycocyanin
(Miller et al. 2013). This is a useful approach for targeted studies,
but phycocyanin is also not always available and measures the
predominance of cyanobacterial pigments and not toxins. Using our
chlorophyll \emph{a} cutoffs provides a value that is more directly
associated with microcystin-LR and can be used to classify lakes, from
past surveys, as having bloomed.

>>>>

Over the last decade or so numerous events and legislative activities
have raised the public awareness of harmful algal blooms (Rinta-Kanto et
al. 2009, HABHRCA 2014, Jetoo et al. 2015), and the US Environmental
Protection Agency (USEPA) has recently released suggested microcystin-LR
(one of the more common toxins) concentrations that would trigger health
advisories (McElhiney and Lawton 2005, Zurawell et al. 2005, USEPA
2015). Additionally, the World Health Organization (WHO) has had
proposed advisory levels for drinking water and a range of recreational
risk levels (Chorus and Bartram 1999, (WHO) and others 2003). While
these levels and associated advisories are likely to help mitigate the
impacts from harmful algal blooms, they are not without complications.

One of these complications is that they rely on available measurements
of microcystin-LR. While laboratory testing remains the gold standard
for quantifying microcystin-LR concentrations in water samples, several
field test kits have been developed. Even though field tests provide a
much needed means for rapid assessment, they are not yet widely used and
are moderately expensive (approximately \$150-\$200 depending on
specific kit) with a limited shelf life (typically one year) (James et
al. 2011, Aranda-Rodriguez et al. 2015). Additionally, each technique
requires nuanced understanding of the detection method (e.g., limit of
detection, specific microcystin variants being measured, and sampling
protocol). Fortunately, microcystin-LR has been shown to be associated
with several other, more commonly measured and understood components of
water quality.

Chlorophyll \emph{a} is a very commonly measured component of water
quality that is also known to be associated with Microsystin-LR
concentrations {[}Lee et al. (2000); Paerl and Otten (2013);
heisler2008eutrophication{]}. Additionally there are many rapid
measurements for assessing chlorophyll \emph{a} levels \emph{in situ}.
For instance, there are small or hand held flourometers that measure
chlorohpyll \emph{a}. Given these facts, it might be possible to
identify chlorophyll \emph{a} concentrations that would be associated
with the various Microcystin-LR health advisory levels. Identifying
these associations would provide another tool for water resource
managers to use to help manage the threat to public health posed by
cHABs and would be especially useful in the absence of microcystin-LR
concentrations. Thus, the goal of this paper is to utilize the National
Lakes Assessment data and identify chlorophyll \emph{a} concentrations
that are associated with higher probabilities of exceeding several
microcystin-LR health advisory concentrations (USEPA 2009). So that
others may repeat or adjust this analysis, the data, code, and this
manuscript are freely available via
\href{https://github.com/USAPE/microcystinchla}{\url{https://github.com/USAPE/microcystinchla}}.

@@

A detailed discussion of CPA is beyond the scope of this paper, but see
Paul et al. (2005) and Hollister et al. (2008) for details. For this
analysis, we used CPA to examine how the conditional probability of
exceeding one of the health advisory changes as chlorophyll \emph{a}
increases in a lake. The 95\% confidence intervals were calculated from
1000 bootstrapped samples. To identify chlorophyll \emph{a}
concentrations of concern we used a 50\% conditional probability of
exceeding each health advisory level and extracted the minimum
chlorophyll \emph{a} concentration that was associated with an upper
confidence level of 50\% or greater. As both microcystin-LR and
chlorophyll \emph{a} values were highely skewed right, a log base 10
transformation was used. Additional details of the specific
implementation are available at
\url{https://github.com/USEPA/microcystinchla}.

@@

In the 2007 NLA, microcystin-LR concentrations ranged from 0.05 to 225.
Microcystin-LR concentrations of 0.05 \(\mu\)g/L represent the detection
limits. Any value greater than that indicates the presence of
microcystin-LR. Of those lakes with microcystin, the median
concentration was 0.51 and the mean was 3.1661047. Lastly, of all lakes
sampled, 21\% of lakes exceeded the U.S. EPA childrens drinking water
standard, 8.8\% of lakes exceeded the U.S. EPA adult drinking water
standard,11.7\% of lakes exceeded the WHO drinking water standard,and
7.3\% of lakes exceeded the WHO recreational standard. For chlorophyll
\emph{a}, the range was 0.07 to 936 \(\mu\)g/L. All lakes had reported
chlorophyll \emph{a} concentrations that exceeded detection limits. The
median concentration was 7.79 and the mean was 29.63. The associations
between chlorophyll \emph{a} and the upper confidence interval with a
conditional probability of 50\% were 22.25, 65.6, 84.96, and 105.84 for
0.3, 1.0, 1.6 and 2.0 \(\mu\)g/L, respectively (Table
\ref{tab:mc_chla_table} \& Figure \ref{fig:multi_cp_plot}).

@@

There are numerous possible uses for the chlorophyll \emph{a} and
microcystin-LR advisory cut-off values. First, in the absence of
microcystin-LR measurements, exceedence of the chlorophyll \emph{a}
concentrations could be a trigger for further actions. Given that there
is uncertainity around these chlorophyll \emph{a} cutoffs the best case
scenario would be to monitor for chlorophyll and in the event of
exceeding a target concentration take water samples and have those
samples tested in a lab for microcystin-LR.

A second potential use is to identify possible bloom events from
historical data. As harmful algal blooms are made up of many species and
have various mechanisms responsible for adverse impacts (e.g.~toxins,
hypoxia, odors), there is no single definition of a bloom. For cHABs one
approach has been to identify an increase over a baseline concentration
of phycocyanin (Miller et al. 2013). This is a useful approach for
targeted studies, but phycocyanin is also not always available and
measures the predominance of cyanobacterial pigments and not toxins.
Using our chlorophyll \emph{a} cutoffs provides a value that is more
directly associated with microcystin-LR and can be used to classify
lakes, from past surveys, as having bloomed.
